<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, eval=TRUE}
opts_knit$set(
  upload.fun = function(file) imgur_upload(file, "ce3138fd1186b7d"), 
  base.url = NULL) # upload all images to imgur.com
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
runchunks = FALSE
saveout   = runchunks

if (!runchunks) load('.README.RData')
```
SpiecEasi
=========

Sparse InversE Covariance estimation for Ecological Association and Statistical Inference

This package will be useful to anybody who wants to infer graphical models for all sorts of compositional data, though primarily intended for microbiome relative abundance data (generated from 16S amplicon sequence data). It also includes a generator for [overdispersed, zero inflated] multivariate, correlated count data. Please see the paper published in [PLoS Comp Bio](http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004226).

One small point on notation: we refer to the method as "SPIEC-EASI" and reserve "SpiecEasi" for this package.

## Installation ##

I assume that all auxillary packages are already installed - for example huge, MASS, etc. If you get an unexpected error, you may need to download and install a missing dependency.

From an interactive R session:

```{r, eval=FALSE}
library(devtools)
install_github("zdk123/SpiecEasi")
library(SpiecEasi)
```
## Basic Usage ##

Lets simulate some multivariate data under zero-inflated negative binomial model, based on (high depth/count) round 1 of the American gut project, with a sparse network. The basic steps are

1. load the data and normalize counts to to common scale (min depth)
2. fit count margins to the model
3. generate a synthetic network
4. generate some synthetic data
5. clr transformation
6. inverse covariance estimation
7. stability selection
8. evaluate performance

Obviously, for real data, skip 1-4.

```{r, eval=runchunks}
data(amgut1.filt)
depths <- rowSums(amgut1.filt)
amgut1.filt.n <- t(apply(amgut1.filt, 1, norm_to_total))
amgut1.filt.cs <- round(amgut1.filt.n * min(depths))

d <- ncol(amgut1.filt.cs)
n <- nrow(amgut1.filt.cs)
e <- d
```
Synthesize the data
```{r, eval=runchunks}
set.seed(10010)
graph <- make_graph('cluster', d, e)
Prec  <- graph2prec(graph)
Cor   <- cov2cor(prec2cov(Prec))

X <- synth_comm_from_counts(amgut1.filt.cs, mar=2, distr='zinegbin', Sigma=Cor, n=n)
```

the main SPIEC-EASI pipeline: Data transformation, sparse invserse covariance estimation and model selection
```{r, eval=runchunks}
se.est <- spiec.easi(X, method='mb', lambda.min.ratio=1e-2, nlambda=15)
```

examine ROC over lambda path and PR over the stars index for the selected graph
```{r, eval=FALSE, fig.width=5, fig.height=5}
huge::huge.roc(se.est$path, graph, verbose=FALSE)
stars.pr(getOptMerge(se.est), graph, verbose=FALSE)
# stars selected final network under: se.est$refit
```

The above example does not cover all possible options and parameters. For example, other generative network models are available, the lambda.min.ratio (the scaling factor that determines the minimum sparsity/lambda parameter) shown here might not be right for your dataset, and its possible that you'll want more repetitions for stars selection.


## Analysis of American Gut data ##


Now let's apply SpiecEasi directly to the American Gut data. Don't forget that the normalization is performed internally in the \lstinline[basicstyle=\ttfamily]|spiec.easi| function. Also, we should use a larger number of stars repetitions for real data. We can pass in arguments to the inner stars selection function as a list via the parameter \lstinline[basicstyle=\ttfamily]|icov.select.params|. If you have more than one processor available, you can also supply a number to \lstinline[basicstyle=\ttfamily]|ncores|. Also, let's compare results from the MB and glasso methods as well as SparCC (correlation).

```{r, eval=runchunks, message=FALSE, warning=FALSE}
se.mb.amgut <- spiec.easi(amgut1.filt, method='mb', lambda.min.ratio=1e-2, 
                            nlambda=20, icov.select.params=list(rep.num=50))
se.gl.amgut <- spiec.easi(amgut1.filt, method='glasso', lambda.min.ratio=1e-2,
                            nlambda=20, icov.select.params=list(rep.num=50))
sparcc.amgut <- sparcc(amgut1.filt)
## Define arbitrary threshold for SparCC correlation matrix for the graph
sparcc.graph <- abs(sparcc.amgut$Cor) >= 0.3
diag(sparcc.graph) <- 0
sparcc.graph <- Matrix(sparcc.graph, sparse=TRUE)
## Create igraph objects
ig.mb <- adj2igraph(se.mb.amgut$refit)
ig.gl <- adj2igraph(se.gl.amgut$refit)
ig.sparcc <- adj2igraph(sparcc.graph)
```

Visualize using igraph plotting:
```{r, eval=TRUE, fig.width=15, fig.height=7, message=FALSE}
library(igraph)
## set size of vertex proportional to clr-mean
vsize <- rowMeans(clr(amgut1.filt, 1))+6
am.coord <- layout.fruchterman.reingold(ig.mb)

par(mfrow=c(1,3))
plot(ig.mb, layout=am.coord, vertex.size=vsize, vertex.label=NA, main="MB")
plot(ig.gl, layout=am.coord, vertex.size=vsize, vertex.label=NA, main="glasso")
plot(ig.sparcc, layout=am.coord, vertex.size=vsize, vertex.label=NA, main="sparcc")
```

We can evaluate the weights on edges networks using the terms from the underlying model. SparCC correlations
can be used directly, while SpiecEasi networks need to be massaged a bit. Note that since SPIEC-EASI
is based on penalized estimators, the edge weights are not directly comparable to SparCC (or Pearson/Spearman
correlation coefficients)

```{r, eval=TRUE, fig.width=8, fig.height=5}
library(Matrix)
elist.gl <- summary(triu(cov2cor(se.gl.amgut$opt.cov)*se.gl.amgut$refit, k=1))
elist.mb <- summary(symBeta(getOptBeta(se.mb.amgut), mode='maxabs'))
elist.sparcc <- summary(sparcc.graph*sparcc.amgut$Cor)

hist(elist.sparcc[,3], main="", xlab="edge weights")
hist(elist.mb[,3], add=TRUE, col='forestgreen')
hist(elist.gl[,3], add=TRUE, col='red')
```

Lets look at the degree statistics from the networks inferred by each method.

```{r, eval=TRUE, fig.width=9, fig.height=6}
dd.gl <- degree.distribution(ig.gl)
dd.mb <- degree.distribution(ig.mb)
dd.sparcc <- degree.distribution(ig.sparcc)

plot(0:(length(dd.sparcc)-1), dd.sparcc, ylim=c(0,.35), type='b', 
      ylab="Frequency", xlab="Degree", main="Degree Distributions")
points(0:(length(dd.gl)-1), dd.gl, col="red" , type='b')
points(0:(length(dd.mb)-1), dd.mb, col="forestgreen", type='b')
legend("topright", c("MB", "glasso", "sparcc"), 
        col=c("forestgreen", "red", "black"), pch=1, lty=1)
```


## Working with phyloseq ##

SpiecEasi includes some convience wrappers to work directly with `phyloseq` objects.
```{r, eval=TRUE, warning=FALSE, message=FALSE}
library(phyloseq)
## Load round 2 of American gut project
data('amgut2.filt.phy')
se.mb.amgut2 <- spiec.easi(amgut2.filt.phy, method='mb', lambda.min.ratio=1e-2,
                           nlambda=20, icov.select.params=list(rep.num=50))
ig2.mb <- adj2igraph(se.mb.amgut2$refit,  vertex.attr=list(name=taxa_names(amgut2.filt.phy)))
plot_network(ig2.mb, amgut2.filt.phy, type='taxa', color="Rank3")
```


```{r, echo = FALSE, eval=TRUE}
if (saveout) 
 save(amgut1.filt, depths, amgut1.filt.n, amgut1.filt.cs, d, n, e, graph, Prec, Cor, se.est, se.mb.amgut, se.gl.amgut, sparcc.amgut, sparcc.graph, ig.mb, ig.gl, ig.sparcc, file=".README.RData")
```
