---
title: "Learning latent variable graphical models"
author: 
  - name: Zachary Kurtz
    email: zdkurtz@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r format(Sys.time(), '%B %d, %Y')`"
package: "SpiecEasi"
version: "1.99.5"
vignette: >
  %\VignetteIndexEntry{Learning latent variable graphical models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, echo = FALSE, eval=TRUE, message=FALSE}
library(BiocStyle)
knitr::opts_knit$set(
  upload.fun = NULL,
  base.url = NULL) # use local files for images
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
# Override BiocStyle plot hook to avoid css_align issues
knitr::knit_hooks$set(plot = function(x, options) {
  paste0('![', basename(x), '](', x, ')')
})
runchunks = if (Sys.getenv("FORCE_VIGNETTE_REBUILD", "FALSE") == "TRUE") TRUE else FALSE

cache_file <- '../vignette_cache/latent-variable-models.RData'
if (!runchunks && file.exists(cache_file)) {
  load(cache_file)
  # If we loaded trimmed objects, reassign them to original names
  if (exists("se1.mb.trimmed")) {
    se1.mb.amgut <- se1.mb.trimmed
    rm(se1.mb.trimmed)
  }
  if (exists("se2.mb.trimmed")) {
    se2.mb.amgut <- se2.mb.trimmed
    rm(se2.mb.trimmed)
  }
  if (exists("se1.slr.trimmed")) {
    se1.slr.amgut <- se1.slr.trimmed
    rm(se1.slr.trimmed)
  }
  if (exists("se2.slr.trimmed")) {
    se2.slr.amgut <- se2.slr.trimmed
    rm(se2.slr.trimmed)
  }
} else {
  if (!runchunks) {
    message("Cache file latent-variable-models.RData not found - building from scratch")
  }
  runchunks <- TRUE
}
saveout   = runchunks
```

# Learning latent variable graphical models

It can be shown that unobserved, latent variables introduce artifacts into empirical estimates of OTU-OTU associations. These effects can be removed from the network by treating the inverse covariance selection problem as a sparse + low-rank decomposition (SPIEC-EASI slr), where the sparse component are the associations encoded as a conditional independence graph, and the low-rank components are the isolated latent effects.

Please see the [preprint](https://www.biorxiv.org/content/10.1101/2019.12.21.885889v1.full) and the manuscript [Synapse project](https://www.synapse.org/#!Synapse:syn20843558) or [github repository](https://github.com/zdk123/SpiecEasiSLR_manuscript) for more details.

To demonstrate this in action we can show that removing latent effects improves a consistency measure between round 1 and round 2 of the American Gut project data.

```{r, eval=TRUE}
library(SpiecEasi)
library(phyloseq)
data(hmp2)
data(amgut1.filt)
data(amgut2.filt.phy)
```

First we fit the networks, assuming that there are 10 latent components in the dataset:

```{r, eval=runchunks}
se1.mb.amgut <- spiec.easi(amgut1.filt, method='mb', lambda.min.ratio=1e-2,
                          nlambda=20, pulsar.params=list(rep.num=20, ncores=4))
se2.mb.amgut <- spiec.easi(amgut2.filt.phy, method='mb', lambda.min.ratio=1e-2,
                          nlambda=20, pulsar.params=list(rep.num=20, ncores=4))

se1.slr.amgut <- spiec.easi(amgut1.filt, method='slr', r=10, lambda.min.ratio=1e-2,
                          nlambda=20, pulsar.params=list(rep.num=20, ncores=4))
se2.slr.amgut <- spiec.easi(amgut2.filt.phy, method='slr', r=10, lambda.min.ratio=1e-2,
                          nlambda=20, pulsar.params=list(rep.num=20, ncores=4))
```

Then we compare the consistency between the edge sets within each method using the Jaccard index:

```{r, eval=TRUE}
otu1 <- colnames(amgut1.filt)
otu2 <- taxa_names(amgut2.filt.phy)
edge.diss(getRefit(se1.mb.amgut), getRefit(se2.mb.amgut), 'jaccard', otu1, otu2)
edge.diss(getRefit(se1.slr.amgut), getRefit(se2.slr.amgut), 'jaccard', otu1, otu2)
```

Consistency should be a bit better for the slr networks.

Construct the robust PCA from amgut2 data:

```{r, eval=TRUE}
# Use cached X and L if available, otherwise extract from objects
if (!exists("X") || !exists("L")) {
  X <- se2.slr.amgut$est$data
  L <- se2.slr.amgut$est$resid[[getOptInd(se2.slr.amgut)]]
}
pca <- robustPCA(X, L)
```

We can also check the correlation between AGP meta-data and the latent factors (scores of the robust PCA):

```{r, eval=TRUE}
age <- as.numeric(as.character(sample_data(amgut2.filt.phy)$AGE))
bmi <- as.numeric(as.character(sample_data(amgut2.filt.phy)$BMI))
depth <- colSums(otu_table(amgut2.filt.phy))

cor(age, pca$scores, use='pairwise')
cor(bmi, pca$scores, use='pairwise')
cor(depth, pca$scores, use='pairwise')
```

## Key differences from standard SPIEC-EASI

The `slr` method in SpiecEasi implements the sparse + low-rank decomposition approach, which:

1. **Removes latent effects**: By decomposing the inverse covariance into sparse and low-rank components, we can isolate and remove the effects of unobserved variables.

2. **Improves consistency**: Networks inferred with latent variable correction tend to be more consistent across different datasets or time points.

3. **Parameter `r`**: Specifies the number of latent components to model. This should be chosen based on domain knowledge or cross-validation.

4. **Computational considerations**: The slr method is more computationally intensive than standard methods, so consider using parallel processing options.

Session info:
```{r}
sessionInfo()
```

```{r, echo = FALSE, eval=TRUE, message=FALSE}
# Save objects if they exist
if (exists("se1.mb.amgut") && exists("se2.mb.amgut")) {
  cache_file <- '../vignette_cache/latent-variable-models.RData'
  tryCatch({
    # Load trimming function and trim objects to reduce size
    source('../vignette_cache/trim_spiec_easi.R')
    se1.mb.trimmed <- trim_spiec_easi(se1.mb.amgut)
    se2.mb.trimmed <- trim_spiec_easi(se2.mb.amgut)
    se1.slr.trimmed <- trim_spiec_easi(se1.slr.amgut)
    se2.slr.trimmed <- trim_spiec_easi(se2.slr.amgut)
    
    # Save trimmed objects and other essential data
    save(se1.mb.trimmed, se2.mb.trimmed, se1.slr.trimmed, se2.slr.trimmed, otu1, otu2, X, L, pca, age, bmi, depth, file=cache_file)
    message("Saved trimmed cache file: ", cache_file, " in directory: ", getwd())
  }, error = function(e) {
    message("Failed to save cache file: ", e$message)
  })
}
``` 