---
title: "Troubleshooting"
author: 
  - name: Zachary Kurtz
    email: zdkurtz@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r format(Sys.time(), '%B %d, %Y')`"
package: "`r pkg_ver('SpiecEasi')`"
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, echo = FALSE, eval=TRUE}
knitr::opts_knit$set(
  upload.fun = function(file) knitr::imgur_upload(file, "ce3138fd1186b7d"),
  base.url = NULL) # upload all images to imgur.com
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
runchunks = TRUE
saveout   = runchunks

if (!runchunks) load('.VIGNETTE.RData')
```

# Troubleshooting

A common issue that comes up when running `spiec.easi` is coming up with an empty network after running StARS.

For example:

```{r, echo=TRUE, message=FALSE, warning=TRUE}
library(SpiecEasi)
data(amgut1.filt)

pargs <- list(seed=10010)
se <- spiec.easi(amgut1.filt, method='mb', lambda.min.ratio=5e-1, nlambda=10, pulsar.params=pargs)
getOptInd(se)
sum(getRefit(se))/2
```

As the warning indicates, the network stability could not be determined from the lambda path. Looking at the stability along the lambda path, `se$select$stars$summary`, we can see that the maximum value of the StARS summary statistic never crosses the default threshold (0.05).

This problem we can fix by lowering `lambda.min.ratio` to explore denser networks:

```{r, message=FALSE, warning=FALSE}
se <- spiec.easi(amgut1.filt, method='mb', lambda.min.ratio=1e-1, nlambda=10, pulsar.params=pargs)
```

We have now fit a network, but since we have only a rough, discrete sampling of networks along the lambda path, we should check how far we are from the target stability threshold (0.05):

```{r}
getStability(se)
sum(getRefit(se))/2
```

To get closer to the mark, we should bump up `nlambda` to more finely sample of the lambda path, which gives a denser network:

```{r, message=FALSE, warning=FALSE}
se <- spiec.easi(amgut1.filt, method='mb', lambda.min.ratio=1e-1, nlambda=100, pulsar.params=pargs)
getStability(se)
sum(getRefit(se))/2
```

## Common issues and solutions

### 1. Empty networks

**Problem**: After running `spiec.easi`, you get an empty network (no edges).

**Solutions**:
- Lower `lambda.min.ratio` to explore denser networks
- Increase `nlambda` for finer sampling of the lambda path
- Check if your data has sufficient signal-to-noise ratio
- Try different methods ('mb' vs 'glasso')

### 2. Very dense networks

**Problem**: The inferred network has too many edges.

**Solutions**:
- Increase `lambda.min.ratio` to explore sparser networks
- Adjust the StARS threshold in `pulsar.params`
- Use cross-validation instead of StARS

### 3. Computational issues

**Problem**: The analysis takes too long or runs out of memory.

**Solutions**:
- Use parallel processing with `ncores` parameter
- Use B-StARS method for large datasets
- Reduce `rep.num` in pulsar.params
- Use batch mode for HPC clusters

### 4. Convergence issues

**Problem**: The algorithm doesn't converge or gives warnings.

**Solutions**:
- Check data preprocessing and normalization
- Ensure data doesn't have constant columns
- Try different starting values
- Check for missing or infinite values

### 5. Memory issues

**Problem**: R runs out of memory during analysis.

**Solutions**:
- Use sparse matrices where possible
- Reduce dataset size by filtering rare taxa
- Use batch processing for large datasets
- Increase system memory if available

## Diagnostic functions

SpiecEasi provides several functions to help diagnose issues:

```{r, eval=FALSE}
# Check stability along lambda path
getStability(se)

# Get optimal lambda index
getOptInd(se)

# Get summary statistics
se$select$stars$summary

# Check network density
sum(getRefit(se))/2

# Visualize stability curve
plot(se$select$stars$summary)
```

## Parameter tuning guidelines

### For small datasets (< 100 samples, < 50 taxa):
- `lambda.min.ratio = 1e-2`
- `nlambda = 20-50`
- `rep.num = 20-50`

### For medium datasets (100-1000 samples, 50-200 taxa):
- `lambda.min.ratio = 1e-3`
- `nlambda = 50-100`
- `rep.num = 50-100`
- Use parallel processing

### For large datasets (> 1000 samples, > 200 taxa):
- `lambda.min.ratio = 1e-4`
- `nlambda = 100+`
- `rep.num = 100+`
- Use B-StARS method
- Consider batch processing

Session info:
```{r}
sessionInfo()
```

```{r, echo = FALSE, eval=runchunks}
if (saveout)
 save(se, pargs, file=".VIGNETTE.RData")
``` 